### 2015-11-12 paper.Rmd
barebones paper with holes for various things
see issues for what to do next

**Need to include more history here (TRANSECT etc)?**

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication. An example of this output is shown in **Table ???** and was generated by:
```{r summary-table-null, eval=FALSE}
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

```{r summary-table, results="asis", echo=FALSE}
kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error. Models are sorted according to AIC.")
```

@@

**stratification**

@@

Model uncertainty can be addressed using standard maximum likelihood theory. We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix. We can then pre- and post-multiple this by the derivatives of $\hat{N}_\text{C}$ 
$$
\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
$$
where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability [@Buckland:2001vm, p 62]. Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$.

@@

The \code{varn} manual page gives further advice and technical detail on this matter.

@@

Going back to the minke whale data, we have the information we require to calculate $A$ and $a$ above, so we can estimate abundance and its variance. When we supply data to \code{ds} in the ``flat file'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
```{r minke-N-summary}
summary(minke_hn)
```

@@

* Interface to dsm etc
* MRDS?
* why is + important?

@@

* Other software? \pkg{unmarked}, \pkg{RDistance} etc

>>>>

@@

Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models. \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once. The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication. An example of this output is shown in **Table ???** and was generated by:
```{r summary-table-null, eval=FALSE}
library(knitr)
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
```

```{r summary-table, results="asis", echo=FALSE}
library(knitr)
kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error. Models are sorted according to AIC.")
```

@@

### Post-stratification

We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*. A stratum may be defined by, for example, a different habitat type or by the sex of the animal (or some combination) which may be of interest for biological or management reasons. In order to calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs. wetland habitat the areas of those strata would be different). Finally, we must also know which stratum a given transect lies in.

As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively. Figure \ref{minke-strata} shows the two strata, along with observations and transect lines.

\begin{figure}
\includegraphics{minke-strata}
\label{minke-strata}
\caption{Strata used for the minke whale data adapted from @Hedley:2004et. Points show the locations of observations along transect lines. The stepped line shows the boundary between North and South strata. Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
\end{figure}

@@

Model uncertainty can be addressed using standard maximum likelihood theory. We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix. We can then pre- and post-multiply this by the derivatives of $\hat{N}_\text{C}$ 
$$
\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
$$
where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability [@Buckland:2001vm, p 62]. Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$.

@@

The \code{varn} manual page gives further advice and technical detail on encounter rate variance.

@@

Going back to the minke whale data, we have the information we require to calculate $A$ and $a$ above, so we can estimate abundance and its variance. When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.

Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
```{r minke-N-summary}
summary(minke_hn)
```
This prints a rather large amount of information. One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication. For example **Table ???**.

```{r minke-abundance-table}
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
                           "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits = 3, format = "latex",
      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.")
```

@@

The features of \pkg{Distance} are delibarately limited to provide a simplified interface for users. However, for more complex distance sampling-based analyses there are further related packages for modelling in \proglang{R}.

We noted at the start of the article that \pkg{Distance} is a simpler wrapper around the package \pkg{mrds}. Additional features are available in \pkg{mrds} including the ability to model data where the assumption that detection is certain at zero distance from the line or point by using mark-recapture type methods when two observers are used [see @Burt:2014gu for an introduction].

The abundance estimates calculated here are based on the assumption that within a given stratum abundance is constant. We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space. A more rigourous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products). \pkg{Distance} interfaces with once such package to perform this type of analysis: \pkg{dsm}. So-called "density surface modelling" uses the generalized additive model framework [e.g. @Wood:2006vg] to build models of abundance (that take into account detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].

@@

Here we have given an introduction as to how to perform a distance sampling analysis in \proglang{R}. We have covered the possible models for detectability, model checking and selection and finally abundance and variance estimation.

In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists. We hope that this paper (itself written in RMarkdown) will provide a useful set of examples for those wishing to persue this.

We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete. **Appendix ???** gives a feature comparison between \pkg{Distance} and the other packages available.

# Acknowledgements

The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and DISTANCE who have contributed bug reports and suggestions for improvements over the years.
